Working with Daily Climate Model Output Data in R and the \pkg{futureheatwaves} Package

### 2017-01-05 paper_drafts/RJournalRevisions/ToSubmit/anderson-eason-barnes.tex
Update paper draft

% !TeX root = RJwrapper.tex
\title{Working with daily climate model output data in R and the
\pkg{futureheatwaves} package}
\author{by G. Brooke Anderson, Colin Eason, Elizabeth A. Barnes}

@@

\abstract{%
Research on climate change impacts can require extensive processing,
especially when using ensemble techniques to incorporate output from
different climate models and different simulations of each model. This
processing can be particularly extensive when identifying and
characterizing multi-day extreme events like heat waves and frost day
spells, as these must be processed from model output with daily time
steps. Further, climate model output is in a format and follows
standards that may be unfamiliar to most R users. Here, we provide an
overview of working with daily climate model output data in R. We then
present the \pkg{futureheatwaves} package, which we developed to ease
the process of identifying, characterizing, and exploring multi-day
extreme events in climate model output.
}

@@

A key source of data for ensemble techniques is the Coupled Model
Intercomparison Project, phase 5 (CMIP5; \citet{taylor2012overview}).
This project brought together dozens of major climate modeling groups
around the world to simulate the same future radiative forcing
scenarios, but with their own models. This created an ensemble of
state-of-the-art climate model projections that allow researchers to
study projections and their uncertainties. Most of these modeling groups
additionally performed more than one simulation for each scenario and
model (i.e.~multiple ensemble members), perturbing the initial
conditions by a very tiny amount to quantify uncertainties due to
internal climate variability.

@@

After this overview, we present the \pkg{futureheatwaves} package, which
we created to aid in identifying and characterizing any type of
multi-day extreme event from daily climate model output (Table
\ref{tab:goals}). The impacts of multi-day extreme events must be
assessed using output in daily time step, unlike other climate impacts
that can be assessed using climate model output at monthly, seasonal, or
yearly time steps. Further, extreme events are identified based on
conditions that are rare for a certain location (e.g., 98th percentile
of local temperature distribution for identifying heat waves)
\citep{IPCCch1}. In this case, assessment is further complicated, as the
event definition must be determined at each study location from climate
model output. Finally, it is often of interest to create summaries of
multiple characteristics of these extreme events. For example, one may
be interested in determining whether the frequency or characteristics
(e.g., length, intensity) of heat waves or warm spells will change under
certain climate change scenarios \citep{IPCCch1}.

The \pkg{futureheatwaves} package handles these challenges and can be
used to identify and characterize a variety of multi-day extreme events
across different ensemble members of one or more climate models. It also
provides some functionality particularly useful in identifying and
characterizing heat waves specifically. Quantification of the impacts of
heat waves on human health suffer from additional sources of uncertainty
beyond those inherent in projections of regional changes in surface
temperature. These include: (1) uncertainty in the definition of a heat
wave itself and (2) uncertainty in the ability of communities to adapt
to changing temperatures (the adaptation scenario). This package
therefore allows the user to create and use a custom extreme event
definition to identify events in the climate model output, as well as
options to explore different scenarios of adaptation to heat.

\begin{table}[t]
\begin{center}
\newcolumntype{L}[1]{>{\raggedright\arraybackslash}p{#1}}
\begin{tabular}{lL{\dimexpr0.8\textwidth-2\tabcolsep\relax}}
\toprule
 & Design goals of the \pkg{futureheatwaves} package \\
\midrule
1 & Make processing of large ensembles of climate simulations more practical for researchers exploring the potential impacts of heat waves and other multi-day extreme events.\\
2 & Speed up processing time by incorporating C++ in event identification.\\
3 & Keep track of the names of climate models and number of ensemble members processed for each.\\
4 & Not only identify, but also characterize, all extreme events within each climate simulation, to allow the exploration of patterns in these characteristics across different simulations and also to allow the use of more complex impacts models, including models that incorporate event characteristics (e.g., event length, event intensity). For example, this package allows the user to apply a health effects model where risk of mortality is not the same for every heat wave, but rather is modified by heat wave length, intensity, or other measured characteristics.\\
5 & Give users extensive power in customizing the process, including allowing custom extreme event definitions.\\
6 & Allow users to easily explore the extreme events identified within all climate simulations.\\
7 & Create output that is in a ``tidy" data format, allowing it to work well with \pkg{ggplot2} for visualization.\\
\bottomrule
\end{tabular}
\end{center}
\caption{Design goals for the \pkg{futureheatwaves} package.}
\label{tab:goals}
\end{table}

@@

For climate impact studies, a main source for climate model output is
the Coupled Model Intercomparison Project, which is currently in its
fifth phase (CMIP5). Over 20 climate modeling groups created one or more
climate models which, for this project, were run using standardized
scenarios \citep{taylor2012overview}. The resulting output is uniform
across modeling groups and has a consistent structure, which allows
comparison of simulations from different models \citep{IPCCch9}. CMIP5
climate model output is archived at a number of different time steps
(e.g., daily, monthly, seasonal, yearly) \citep{taylor2010cmip5}, and
some variables are reported at multiple levels in the ocean or
atmosphere (e.g., ocean temperature). Here, we will focus on data with a
daily time step for variables reported at a single level (e.g.,
near-surface air temperature).

Each modeling group ran simulations under several experiments, with
experiments varying in terms of radiative forcing by using different
scenarios of time-varying model inputs (greenhouse gas emissions or
concentrations, land use changes, etc.)
\citep{taylor2012overview, IPCCch9}. Experiments include historical
experiments (run using radiative forcing consistent with observed and
reconstructed data for 1850--2005), pre-Industrial control experiments,
and experiments of future scenarios of radiative forcing over the 21st
century or longer (e.g., RCP4.5, RCP8.5) \citep{taylor2012overview}.
Some modeling groups created ensembles of output for a specific model
and experiment, in which they ran the experiment multiple times with the
model with very small changes to the initial conditions.

@@

CMIP5 files are saved in Network Common Data Format (netCDF), a binary
file format that allows storage of data representing a regular array.
For climate model output at a single level (e.g., near-surface air
temperature), the data is a 3-dimensional array, with dimensions
representing time and two coordinates of location (e.g., latitude and
longitude). Figure \ref{fig:netcdfexample} provides a sketch of the
structure of netCDF files for single-level climate model output. Global
climate models generate output at regularly-spaced time steps, typically
at regularly-spaced grid points around the world. The latitude and
longitude spacing of grid points vary by climate model, but are
typically 1--2 degrees for atmospheric variables in CMIP5 models
\citep{IPCCch9}. Each data point in the netCDF array gives the modeled
value of the variable (e.g., surface temperature) for a single time
point and location. For CMIP5 climate model output, the location units
are in degrees east and degrees north for longitude and latitude,
respectively. For daily output files, the time unit is in days since a
specified origin date-time (e.g., days since 1850-01-01 00:00:00)
\citep{taylor2010cmip5}. All CMIP5 output files are required to include
certain metadata \citep{taylor2010cmip5}, including the experiment,
forcing agents input to the model to create the simulation, time step,
institution and institutional contact information, climate model, and
modeling realm \citep{taylor2010cmip5}. The metadata also must include
units for all of the coordinate variables (e.g., longitude, latitude,
time). The netCDF format allows you to access metadata and variables
describing the dimensions of the data without reading the full file into
memory.

\begin{figure}
\begin{center}
\includegraphics[width = \textwidth]{netcdf_figure}
\end{center}
\caption{Example of structure of a NetCDF climate model output file for a variable reported at a single level, like near-surface air temperature. Data are stored in a three-dimensional array, with measurements at each time step and climate grid location. These data are typically indexed in the netCDF file by longitude, latitude, and time, in that order. For example, if the near-surface air temperature is read into an R object called \code{tas}, you can access the value for the first day at the fourth longitude and third latitude with \code{tas[4, 3, 1]}. In addition to the output variable (temperature in this example), vectors with the ordered values of each dimension (longitude, latitude, and time) can also be read in from the netCDF file, as well as attribute data (e.g., units for variables, the calendar used for time).}
\label{fig:netcdfexample}
\end{figure}

@@

CMIP5 data are available as netCDF files, for which base R import
functions do not exist. There are, however, a few R packages that can be
used to work with the netCDF file format used for CMIP5 files. Older
packages include \pkg{ncdf} and \pkg{ncvar}, but these do not work with
the newer netCDF version 4 released in 2008 and are no longer available
through CRAN. More recent packages, including \pkg{ncdf4} \citep{ncdf4}
and \pkg{RNetCDF} \citep{michna2013rnetcdf, RNetCDF}, work with both
version 4 and netCDF's older version 3. While climate model output data
for CMIP5 are required to conform with the earlier version (version 3)
\citep{taylor2010cmip5}, it is safer to write code using functions that
can be used with either version, in case future phases of CMIP do not
require files to conform with netCDF version 3.

You can do a number of things with netCDF files in R using these
functions. For example, \pkg{ncdf4}'s \code{nc\_open} function can be
used to open a connection to a netCDF file, and the object returned by
the function includes the file's attribute data. Once a file connection
is open, variables can be read in using the \code{ncvar\_get} function.
For example, the variables defining the dimensions of the sketched
netCDF file in Figure \ref{fig:netcdfexample} could be read into R with
\code{ncvar\_get} with the \code{varid} parameter set to ``lat'',
``lon'', or ``time''. The climate output variable (e.g., near-surface
air temperature) can similarly be read in using \code{ncvar\_get}. In
this case, the \code{varid} parameter should be set using the
appropriate CMIP5 variable name (e.g., ``tas'' for near-surface air
temperature); these variable names can be found in the CMIP Requested
Output tables \citep{taylor2010cmip5}. If you only need data from a
subset of the full file, you can use the dimensional time and location
data to identify the location of the variable data you need in the
netCDF array and use indexing to read only that data into memory,
without needing to read in the full file (for example, with the
\code{nc.get.var.subset.by.axes} function in \pkg{ncdf4.helpers}). Once
the user is done reading in data from the file, the connection can be
closed (e.g., with the \code{nc\_close} function from \pkg{ncdf4}).

@@

Two R packages provide help with non-Gregorian calendars: \pkg{PCICt}
\citep{PCICt} and \pkg{ncdf4.helpers} \citep{ncdf4.helpers}. The
\code{nc.get.time.series} function in \pkg{ncdf4.helpers} pulls and uses
metadata on the calendar stored in the CMIP5 netCDF file's attributes to
convert the ``time'' variable in the file to an object of the
\code{PCICt} class. This class is defined in the \pkg{PCICt} package and
provides Date-like functionality for 360- and 365-day calendars
\citep{PCICt}. However, while these functions will help with handling
most CMIP5 files, the CMIP5 standards allows use of other calendars
which may not be succesfully handled by these functions, so it is
important to assess whether the time variable range in the \code{PCICt}
object correctly matches the expected date ranges for a file when
processing CMIP5 data in R.

@@

In addition to these general packages for working with netCDF files,
there are several R packages specifically for working with climate model
output data, including \pkg{RCMIP5} \citep{RCMIP5} and \pkg{wux}
\citep{wux}. However, these packages are more useful for working with
data output at time steps of a month or higher and have limited utility
with the daily climate model output data required for studies of
multi-day extreme events.

The \pkg{RCMIP5} package includes functions to read in CMIP5 data from
netCDF files, scan a directory of CMIP5 files and determine models with
continuous available data, create objects of a special \code{cmip5data}
class to work with CMIP5 data within R, and parse the filenames for all
files in a directory to extract information within the filename. For
this package, most functions only work with monthly or less frequent
data \citep{RCMIP5}. While the \code{loadCMIP5} function does
successfully load daily data as a \code{cmip5data} object, most of the
methods for this object type do not do anything meaningful for daily
data. The package's \code{getFileInfo} function, however, will work with
CMIP5 files of any time step; this function identifies all CMIP5 files
in a directory and creates a data frame with information parsed from the
file name. As a note, the \code{get.split.filename.cmip5} function in
the \pkg{ncdf4.helper} package similarly can be used to parse
information contained in CMIP5 file names \citep{ncdf4.helpers}.

The \pkg{wux} package \citep{wux} includes functions that allow the user
to download CMIP5 monthly-aggregated output directly from within R with
the \code{CMIP5fromESGF} function. The package then uses the
\code{models2wux} function to read in climate model output netCDF files
and convert them to ``WUX'' data frames, which can be used by other
functions in the package. While this function can input climate model
output with daily time steps (the ``what.timesteps'' element of the
\code{modelinput} list input must be set to ``daily''), the function
aggregates this data to a monthly or less frequent (e.g., seasonal)
aggregation when creating the WUX data frame. Therefore, while this
package provides very useful functionality for working with averaged
output of daily climate model output data, it cannot easily be used to
identify and characterize multi-day extreme events like heat waves.

The functions and packages described in this section can be used with
CMIP5 netCDF files to do things in R like map near-surface air
temperatures from a single climate model on a specific day (Figure
\ref{fig:worldmap}) or pull a time series of daily near-surface air
temperature simulations at a specific climate model grid point (Figure
\ref{fig:timeseries}). The ``starting\_from\_netcdf'' vignette that
comes with the ``futureheatwaves'' package provides all code required to
create these figures, as well as more details and code examples on
working with CMIP5 netCDF files in R.

@@

Once the user creates these data frames of location-specific heat waves,
the \code{apply\_all\_models} function can be used to apply custom
functions across all the extreme event data frames generated by
\code{gen\_hw\_set}. This functionality allows users to create summaries
of extreme events across all climate models and ensemble members (Figure
\ref{fig:overview}, right). The function can be used to generate summary
statistics (e.g., determine average heat wave length or total frost
days) or to apply more complex functions (e.g., apply epidemiologic
effect estimates across the heat waves to generate health impact
estimates).

When using this package, CMIP5 climate model output data require some
pre-processing. Data will need to be saved in a specific format, with
files stored in a specific directory structure. Full details of the
required file and directory structure are provided in the package's
``futureheatwaves'' vignette, while tips and an R script for conducting
this processing starting from CMIP5 netCDF files are given in the
``starting\_from\_netcdf'' vignette.

This package can be used for study locations worldwide. The
``starting\_from\_netcdf''" package vignette provides an example of
using this package to identify and explore future heat waves in several
Chinese cities.

@@

We have included data files in the package to serve as example files so
that users can try this package before applying it to their own
directory of climate projection files. These example data come from two
climate models that are a part of CMIP5: (1) the model of the Beijing
Climate Center, China Meteorological Administration (BCC)
\citep{xin2013introduction} and (2) the National Center for Atmospheric
Research's (NCAR's) Community Climate System Model, version 4 (CCSM4)
\citep{gent2011community}. We include one ensemble member from BCC
(r1i1p1) and two from CCSM (r1i1p1 and r2i1p1). Once the
\pkg{futureheatwaves} package is installed and loaded, the user can find
the local location of these files using R's \code{system.file} function.

@@

This code first identifies and saves as objects the path names on the
user's computer of the example climate projections directory
(\code{projection\_dir\_location}) and the file of study locations
(\code{city\_file\_location}). The \code{gen\_hw\_set} function
processes the example input and creates a new directory,
\file{example\_results}, with files of identified and characterized heat
waves, in the user's current working directory. In this example code,
the processing is done using default values for the event definition,
years for which to generate the heat wave data sets, etc. How and why to
customize these choices are explained later in the text. Function
arguments (e.g., \code{dataDirectories}, \code{tasFilenames}) are used
to specify the format of the data and the directory structure.

@@

Once the user has created a directory of characterized event files for
each ensemble member (``Extreme event data sets'', Figure
\ref{fig:overview}), he or she can explore the results using the
\code{apply\_all\_models} function. This function allows the user to
apply custom R functions across all extreme event data frames created by
the \code{gen\_hw\_sets} call. The user can apply any R function that
follows certain standards in accepting input and returning output. Full
details on these standards are given in the \pkg{futureheatwaves}
package vignette.

As an example, if the user wanted to calculate the average temperature
of the heat waves identified for each ensemble member, he or she could
write a simple function:

@@

\noindent This function could then be applied across all extreme event
data sets output by \code{gen\_hw\_set} using the
\code{apply\_all\_models} function. For example, to apply this function
to all the example output results that come with the package, the user
could run:

@@

This output is structured as ``tidy" data \citep{wickham2014tidy},
allowing it to be used easily with the graphing package
\CRANpkg{ggplot2} \citep{ggplot2}. This functionality can be used to
create a number of other summaries of the identified extreme events. For
example, it could be used to determine average heat wave length or
estimate how much earlier in the year events are expected to start
across an ensemble of climate model simulations. The functionality can
also be used for more complex analysis of extreme event files. For
example, it can be used to apply epidemiological models of heat wave to
estimate excess heat-related mortality under different future scenarios;
an example of this application is provided in the ``futureheatwaves''
vignette that comes with the package.

@@

Therefore, this package allows the user to extensively customize the
definition used to identify extreme events. Users can write a custom R
function with either a different heat wave definition (see
\citet{smith2013heat} and \citet{kent2014heat} for listings of some of
the definitions used in scientific studies) or with a definition
appropriate for a different type of extreme event (e.g., one or more
days at or below 32\textsuperscript{o}F / 0\textsuperscript{o}C for
frost day spells). For heat wave identification, researchers might want
to use a specific definition, for example, because it matches the
definition used by local health officials to declare heat wave warnings
or, in the case of health impact assessments, to match with a definition
used in an epidemiological study. For studies of other extreme events, a
heat wave definition likely will not be applicable and so a customized
definition is necessary.

>>>>

% !TeX root = RJwrapper.tex
\title{Working with Daily Climate Model Output Data in R and the
\pkg{futureheatwaves} Package}
\author{by G. Brooke Anderson, Colin Eason, Elizabeth A. Barnes}

@@

\abstract{%
Research on climate change impacts can require extensive processing of
climate model output, especially when using ensemble techniques to
incorporate output from multiple climate models and multiple simulations
of each model. This processing can be particularly extensive when
identifying and characterizing multi-day extreme events like heat waves
and frost day spells, as these must be processed from model output with
daily time steps. Further, climate model output is in a format and
follows standards that may be unfamiliar to most R users. Here, we
provide an overview of working with daily climate model output data in
R. We then present the \CRANpkg{futureheatwaves} package, which we
developed to ease the process of identifying, characterizing, and
exploring multi-day extreme events in climate model output. This package
can input a directory of climate model output files, identify all
extreme events using customizable event definitions, and summarize the
output using user-specified functions.
}

@@

A key source of data for ensemble techniques is the Coupled Model
Intercomparison Project, phase 5 (CMIP5; \citet{taylor2012overview}).
This project brought together major climate modeling groups around the
world to simulate the same future radiative forcing scenarios, but with
their own models. This created an ensemble of state-of-the-art climate
model projections that allow researchers to study projections and their
uncertainties. Most of these modeling groups additionally performed more
than one simulation for each scenario and model (i.e.~multiple ensemble
members), perturbing the initial conditions by a very tiny amount to
quantify uncertainties due to internal climate variability.

@@

After this overview, we present the \CRANpkg{futureheatwaves} package,
which we created to aid in identifying and characterizing any type of
multi-day extreme event from daily climate model output (Table
\ref{tab:goals}). The impacts of multi-day extreme events must be
assessed using output in daily time step, unlike other climate impacts
that can be assessed using climate model output at monthly, seasonal, or
yearly time steps. Further, extreme events are identified based on
conditions that are rare for a certain location (e.g., 98th percentile
of local temperature distribution for identifying heat waves)
\citep{IPCCch1}. In this case, the event definition must be determined
at each study location from climate model output before events can be
identified. Finally, it is often of interest to create summaries of
multiple characteristics of these extreme events. For example, one may
be interested in determining whether the frequency or characteristics
(e.g., length, intensity) of heat waves or warm spells will change under
certain climate change scenarios \citep{IPCCch1}.

The \pkg{futureheatwaves} package handles these challenges and can be
used to identify and characterize a variety of multi-day extreme events
across different ensemble members of one or more climate models. It also
provides some functionality particularly useful in identifying and
characterizing heat waves specifically. Quantification of the impacts of
heat waves on human health suffer from additional sources of uncertainty
beyond those inherent in projections of regional changes in surface
temperature. These include: (1) uncertainty in the definition of a heat
wave itself and (2) uncertainty in the ability of communities to adapt
to changing temperatures (the adaptation scenario). This package
therefore allows the user to create and use a custom extreme event
definition to identify events in the climate model output, as well as
provides options to explore different scenarios of adaptation to heat.

\begin{table}[t]
\begin{center}
\newcolumntype{L}[1]{>{\raggedright\arraybackslash}p{#1}}
\begin{tabular}{lL{\dimexpr0.8\textwidth-2\tabcolsep\relax}}
\toprule
 & Design goals of the \pkg{futureheatwaves} package \\
\midrule
1 & Make processing of large ensembles of climate simulations more practical for researchers exploring the potential impacts of heat waves and other multi-day extreme events.\\
2 & Speed up processing time by incorporating C++ in event identification.\\
3 & Keep track of the names of climate models and number of ensemble members processed for each.\\
4 & Not only identify, but also characterize, all extreme events within each climate simulation, to allow the exploration of patterns in these characteristics across different simulations and also to allow the use of more complex impacts models, including models that incorporate event characteristics (e.g., event length, event intensity). For example, this package allows the user to apply a health effects model where risk of mortality is not the same for every heat wave, but rather is modified by heat wave length, intensity, or other measured characteristics.\\
5 & Give users extensive power in customizing the process, including allowing custom extreme event definitions.\\
6 & Allow users to easily explore the extreme events identified within all climate simulations.\\
7 & Create output that is in a ``tidy" data format, allowing it to work well with \CRANpkg{ggplot2} for visualization.\\
\bottomrule
\end{tabular}
\end{center}
\caption{Design goals for the \pkg{futureheatwaves} package.}
\label{tab:goals}
\end{table}

@@

For climate impact studies, a main source of data is the Coupled Model
Intercomparison Project, which is currently in its fifth phase (CMIP5).
Over 20 climate modeling groups created one or more climate models
which, for this project, were run using standardized scenarios
\citep{taylor2012overview}. The resulting output is uniform across
modeling groups and has a consistent structure, which allows comparison
of simulations from different models \citep{IPCCch9}. CMIP5 climate
model output is archived at a number of different time steps (e.g.,
daily, monthly, seasonal, yearly) \citep{taylor2010cmip5}, and some
variables are reported at multiple levels in the ocean or atmosphere
(e.g., ocean temperature is reported at different ocean depths). Here,
we will focus on data with a daily time step for variables reported at a
single level (e.g., near-surface air temperature).

Each modeling group ran simulations for CMIP5 under several experiments,
with experiments varying in terms of radiative forcing scenarios through
the use of different scenarios of time-varying model inputs (greenhouse
gas emissions or concentrations, land use changes, etc.)
\citep{taylor2012overview, IPCCch9}. Experiments include historical
experiments (run using radiative forcing consistent with observed and
reconstructed data for 1850--2005), pre-Industrial control experiments,
and experiments of future scenarios of radiative forcing over the 21st
century or longer (e.g., RCP4.5, RCP8.5) \citep{taylor2012overview}.
Some modeling groups created ensembles of output for a specific model
and experiment, in which they ran the experiment multiple times with the
model with very small changes to the initial conditions.

@@

CMIP5 files are saved in Network Common Data Format (netCDF), a binary
file format that allows storage of data representing a regular array.
For climate model output at a single level (e.g., near-surface air
temperature), the data is a 3-dimensional array, with dimensions
representing time and two coordinates of location (e.g., latitude and
longitude). Figure \ref{fig:netcdfexample} provides a sketch of the
structure of netCDF files for single-level climate model output.

Each data point in the netCDF array gives the modeled value of the
variable (e.g., surface temperature) for a single time point and
location. Global climate models generate output at regularly-spaced time
steps, typically at regularly-spaced grid points around the world. The
latitude and longitude spacing of grid points vary by climate model, but
are typically 1--2 degrees for atmospheric variables in CMIP5 models
\citep{IPCCch9}. For CMIP5 climate model output, the location units are
in degrees east and degrees north for longitude and latitude,
respectively. For daily output files, the time unit is in days since a
specified origin date-time (e.g., days since 1850-01-01 00:00:00)
\citep{taylor2010cmip5}.

All CMIP5 output files are required to include certain metadata (or
``attributes'') \citep{taylor2010cmip5}, including the experiment,
forcing agents input to the model to create the simulation, time step,
institution and institutional contact information, climate model, and
modeling realm \citep{taylor2010cmip5}. The metadata also must include
units for all of the dimension variables (e.g., longitude, latitude,
time). The netCDF format allows you to access metadata and variables
describing the dimensions of the data without reading the full file into
memory.

\begin{figure}
\begin{center}
\includegraphics[width = \textwidth]{netcdf_figure}
\end{center}
\caption{Example of structure of a netCDF climate model output file for a variable reported at a single level, like near-surface air temperature. Data are stored in a three-dimensional array, with measurements at each time step and climate grid location. These data are typically indexed in the netCDF file by longitude, latitude, and time, in that order. For example, if the near-surface air temperature in the example netCDF shown here is read into an R object called \code{tas}, you can access the value for the first day at the fourth longitude and third latitude with \code{tas[4, 3, 1]}. In addition to the output variable (temperature in this example), vectors with the ordered values of each dimension (longitude, latitude, and time) can also be read in from the netCDF file, as well as attribute data (e.g., units for variables, the calendar used for time).}
\label{fig:netcdfexample}
\end{figure}

@@

CMIP5 data are available as netCDF files, for which base R import
functions do not exist. There are, however, a few R packages that can be
used to work with the netCDF file format used for CMIP5 files. Older
packages include \pkg{ncdf} and \pkg{ncvar}, but these do not work with
the newer netCDF version 4 released in 2008 and are no longer available
through CRAN. More recent packages, including \CRANpkg{ncdf4}
\citep{CRANncdf4} and \pkg{CRANRNetCDF}
\citep{michna2013rnetcdf, RNetCDF}, work with both version 4 and
netCDF's older version 3. While climate model output data for CMIP5 are
required to conform with the earlier version (version 3)
\citep{taylor2010cmip5}, it is safer to write code using functions that
can be used with either version, in case future phases of CMIP do not
require files to conform with netCDF version 3.

You can do a number of things with netCDF files in R using these
packages. For example, \pkg{ncdf4}'s \code{nc\_open} function can be
used to open a connection to a netCDF file, and the object returned by
the function includes the file's attribute data. Once a file connection
is open, variables can be read in using the \code{ncvar\_get} function.
For example, the variables defining the dimensions of the sketched
netCDF file in Figure \ref{fig:netcdfexample} could be read into R with
\code{ncvar\_get} with the \code{varid} parameter set to ``lat'',
``lon'', or ``time''.

The climate output variable (e.g., near-surface air temperature) can
similarly be read in using \code{ncvar\_get}. In this case, the
\code{varid} parameter should be set using the appropriate CMIP5
variable name (e.g., ``tas'' for near-surface air temperature); these
variable names can be found in the CMIP Requested Output tables
\citep{taylor2010cmip5}. If only a subset of the full file is needed,
the dimensional time and location data can be used to identify the
location of the needed data in the netCDF array and this information can
then be used to read a portion of data into memory (for example, with
the \code{nc.get.var.subset.by.axes} function in
\CRANpkg{ncdf4.helpers}). Once the user is done reading in data from the
file, the connection to the netCDF should be closed (e.g., with the
\code{nc\_close} function from \pkg{ncdf4}).

@@

Two R packages provide help with non-Gregorian calendars:
\CRANpkg{PCICt} \citep{PCICt} and \CRANpkg{ncdf4.helpers}
\citep{ncdf4.helpers}. The \code{nc.get.time.series} function in
\pkg{ncdf4.helpers} pulls and uses metadata on the calendar stored in
the CMIP5 netCDF file's attributes to convert the ``time'' variable in
the file to an object of the \code{PCICt} class. This class is defined
in the \pkg{PCICt} package and provides date-like functionality for 360-
and 365-day calendars \citep{PCICt}. However, while these functions will
help with handling most CMIP5 files, the CMIP5 standards allows use of
other calendars which may not be succesfully handled by these functions,
so it is important to assess whether the time variable range in the
\code{PCICt} object correctly matches the expected date ranges for a
file when processing CMIP5 data in R.

@@

In addition to these general packages for working with netCDF files,
there are several R packages specifically for working with climate model
output data, including \CRANpkg{RCMIP5} \citep{RCMIP5} and \CRANpkg{wux}
\citep{wux}. However, these packages are more useful for working with
data output at time steps of a month or higher and have limited utility
with the daily climate model output data required for studies of
multi-day extreme events.

The \pkg{RCMIP5} package includes functions to read in CMIP5 data from
netCDF files, scan a directory of CMIP5 files and determine models with
continuous available data, create objects of a special \code{cmip5data}
class to work with CMIP5 data within R, and parse the filenames for all
files in a directory to extract information within the filename. For
this package, most functions only work with monthly or less frequent
data \citep{RCMIP5}. While the \code{loadCMIP5} function does
successfully load daily data as a \code{cmip5data} object, most of the
methods for this object type do not do anything meaningful for daily
data. The package's \code{getFileInfo} function, however, will work with
CMIP5 files of any time step; this function identifies all CMIP5 files
in a directory and creates a data frame with information parsed from the
file name. The \code{get.split.filename.cmip5} function in the
\pkg{ncdf4.helper} package similarly can be used to parse information
contained in CMIP5 file names \citep{ncdf4.helpers}.

The \pkg{wux} package \citep{wux} includes functions that allow the user
to download CMIP5 output at a monthly time step directly from R with the
\code{CMIP5fromESGF} function. The package then uses the
\code{models2wux} function to read climate model output netCDF files and
convert it to ``WUX'' data frames, which can be used by other functions
in the package. While this function can input climate model output with
daily time steps (the ``what.timesteps'' element of the
\code{modelinput} list input must be set to ``daily''), the function
aggregates this data to a monthly or less frequent (e.g., seasonal)
aggregation when creating the WUX data frame. Therefore, while this
package provides very useful functionality for working with averaged
output of daily climate model output data or with output data at a
larger time step, it cannot easily be used to identify and characterize
multi-day extreme events like heat waves.

The functions and packages described in this section can be used with
CMIP5 netCDF files to do things in R like map near-surface air
temperatures from a single climate model on a specific day (Figure
\ref{fig:worldmap}) or pull a time series of daily near-surface air
temperature simulations at a specific climate model grid point (Figure
\ref{fig:timeseries}). The \pkg{futureheatwaves}' ``Starting from netCDF
files'' vignette
(\url{https://cran.r-project.org/web/packages/futureheatwaves/vignettes/starting_from_netcdf.html})
provides all code required to create these figures, as well as more
details and code examples on working with CMIP5 netCDF files in R.

@@

Once the user creates these data frames of location-specific extreme
events, the \code{apply\_all\_models} function can be used to apply
custom functions across all the extreme event data frames. This
functionality allows users to create summaries of extreme events across
all climate models and ensemble members (Figure \ref{fig:overview},
right). The function can be used to generate summary statistics (e.g.,
determine average heat wave length or total frost days) or to apply more
complex functions (e.g., apply epidemiologic effect estimates across the
heat waves to generate health impact estimates).

When using this package, CMIP5 climate model output data require some
pre-processing. Data will need to be saved in a specific format, with
files stored in a specific directory structure. Full details of the
required file and directory structure are provided in the package's main
vignette
(\url{https://cran.r-project.org/web/packages/futureheatwaves/vignettes/futureheatwaves.html}),
while tips and an R script for conducting this processing starting from
CMIP5 netCDF files are given in the ``Starting from netCDF files''
vignette
(\url{https://cran.r-project.org/web/packages/futureheatwaves/vignettes/starting_from_netcdf.html}).

This package can be used for study locations worldwide. The ``Starting
from netCDF files'' package vignette provides an example of using this
package to identify and explore future heat waves in several Chinese
cities.

@@

We have included data files in the package to serve as example files so
that users can try this package before applying it to their own
directory of climate projection files. These example data come from two
climate models that are a part of CMIP5: (1) the model of the Beijing
Climate Center, China Meteorological Administration (BCC)
\citep{xin2013introduction} and (2) the National Center for Atmospheric
Research's (NCAR's) Community Climate System Model, version 4 (CCSM4)
\citep{gent2011community}. We include one ensemble member from BCC
(r1i1p1) and two from CCSM (r1i1p1 and r2i1p1). Once the
\pkg{futureheatwaves} package is installed and loaded, the user can find
the location of these files on his or her computer using R's
\code{system.file} function.

@@

This code first identifies and saves as objects the path names on the
user's computer of the example climate projections directory
(\code{projection\_dir\_location}) and the file of study locations
(\code{city\_file\_location}). The \code{gen\_hw\_set} function
processes the example input and creates a new directory,
\file{example\_results}, with files of identified and characterized heat
waves, in the user's current working directory. In this example code,
the processing is done using default values for the event definition,
adaptation scenario, etc. How and why to customize these choices are
explained later in the text. Function arguments (e.g.,
\code{dataDirectories}, \code{tasFilenames}) are used to specify the
format of the data and the directory structure.

@@

Once the user has created a directory of characterized event files for
each ensemble member (``Extreme event data sets'', Figure
\ref{fig:overview}), he or she can explore the results using the
\code{apply\_all\_models} function. This function allows the user to
apply custom R functions across all extreme event data frames created by
the \code{gen\_hw\_sets} call. The user can apply any R function that
follows certain standards in accepting input and returning output. Full
details on these standards are given in the main package vignette.

As an example, if the user wanted to calculate the average temperature
of the heat waves identified for each ensemble member in the output
generated by the code above, he or she could write a simple function:

@@

\noindent This function can then be applied across all extreme event
data sets output by \code{gen\_hw\_set} using the
\code{apply\_all\_models} function. For example, to apply this function
to all the example output results that come with the package, the user
could run:

@@

The same process can be used to create a number of other summaries of
the identified extreme events. For example, it could be used to
determine average length of extreme events or estimate how much earlier
in the year events are expected to start across an ensemble of climate
model simulations. The functionality can also be used for more complex
analysis of extreme event files. For example, it can be used to apply
epidemiological models of heat wave to estimate excess heat-related
mortality under different future scenarios; an example of this
application is provided in the main \pkg{futureheatwaves} vignette. The
output from \code{apply\_all\_models} is structured as ``tidy" data
\citep{wickham2014tidy}, allowing it to be used easily with the graphing
package \CRANpkg{ggplot2} \citep{ggplot2} and other packages in the
tidyverse.

@@

Therefore, this package allows the user to extensively customize the
definition used to identify extreme events. Users can write a custom R
function with either a different heat wave definition (see
\citet{smith2013heat} and \citet{kent2014heat} for listings of some of
the definitions used in scientific studies) or with a definition
appropriate for a different type of extreme event (e.g., one or more
days at or below 32\textsuperscript{o}F / 0\textsuperscript{o}C for
frost day spells). For heat wave identification, researchers might want
to use a different event definition because, for example, it matches the
definition used by local health officials to declare heat wave warnings
or, in the case of health impact assessments, to match with a definition
used in an epidemiological study. For studies of other extreme events, a
heat wave definition likely will not be applicable and so a customized
definition is necessary.
